---
title: "Explanatory analysis"
output:
  html_document: default
  pdf_document: default
editor_options: 
  chunk_output_type: console
---
```{r set-options, echo=FALSE, cache=TRUE}
options(width = 1900)
```

```{r zero, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE, eval = TRUE)
```


```{r,echo=FALSE,collapse=TRUE,message=FALSE,warning=FALSE}


rm(list=ls())
setwd("~/Dropbox/Akdemir-Sanchez/JULIO/GWAS_OATS/Project_GWAS")
source("multiplot.R")
packages <- c( "openxlsx","png","gapminder","randomcoloR",
              "BLR","ggpubr","corrplot","RNOmni", "knitr", "ggrepel","EMMREML","STPGA","sommer","stringi","knitr","readxl","stringr","gridExtra","ggfortify","factoextra","FactoMineR","cluster","tables","kableExtra","beeswarm","ggpmisc","egg","rrBLUP","Hmisc","reshape","GLMMadaptive","PerformanceAnalytics","ggpubr","igraph","tidyverse","extrafont","extrafontdb","cowplot","sommer","insight")
package.check <- lapply(packages, FUN = function(x) {
  if (!require(x, character.only = TRUE)) {
    install.packages(x, dependencies = TRUE)
    library(x, character.only = TRUE)
  }
})

load("/Users/User/Dropbox/Akdemir-Sanchez/JULIO/GWAS_OATS/Project_GWAS/Datareadybefore_Gwas.RData")

#load("/Users/jis/Dropbox/Akdemir-Sanchez/JULIO/GWAS_OATS/Project_GWAS/Datareadybefore_Gwas.RData")
```



```{r}

#str(dat)
#sum(complete.cases(dat$origin)=="FALSE")


#dat <- dat[na.omit(dat$origin),]
colnames(pheno)[24] <- "SumT2HT2"
colnames(pheno)[26] <- "T2"
colnames(pheno)[27] <- "HT2"
dat <- pheno
dat <- dat[!is.na(dat$origin), ]
dat$origin <- droplevels(dat$origin,exclude=c("GER"))
dat <- dat[!is.na(dat$origin), ]
levels(dat$origin) <- c("Finland","France","United Kigndom","Hungary","Ireland","Norway","Poland", "Sweden")
datoxin <- dat[,c(1:11,24:35,39)]

#str(dat)
#str(datoxin)
#table(dat$origin)
#table(datoxin$origin)
#library(plyr)
#newlabels <- revalue(dat$origin, c("1"="Finland","2"= "France","3"= "United Kigndom","4"= "Hungary","5"="Ireland","6"= "Norway","7"= "Poland","8"= "Sweden"))
#PercentageH <- datoxin$Sum_transf[datoxin$High_Low=="High"]/sum(datoxin$Sum_transf)
#datoxin$Sum_transf[datoxin$High_Low=="Low"]/sum(datoxin$Sum_transf)
```


```{r}
#font_import()
#loadfonts()



datoxin %>% 
  count(origin,High_Low,Sum_transf,sort=TRUE) %>% 
  mutate(origin= fct_reorder(origin,Sum_transf,sum)) %>% 
  ggplot(aes(origin,Sum_transf,fill=High_Low))+
  geom_bar(stat = "identity")+
  coord_flip()+
  expand_limits(y=0) +
  scale_fill_grey(start = 0.8, end = 0.5)+
  facet_wrap(~ High_Low)+
  labs(x="Country of origin",y="Counts with High or Low Sum of T2_HT2 toxin")+
  theme_classic()+
  theme(legend.position = "none",text=element_text(size=12, family="Times New Roman"))
        #axis.text.y = element_text(color = "grey20", size = 12, angle = 0, hjust = 1, vjust = 0, face = "plain"))

```



```{r}
datoxin %>% 
  count(origin,High_Low,Sum_transf,sort=TRUE) %>% 
  mutate(origin= fct_reorder(origin,Sum_transf,sum)) %>% 
  ggplot(aes(origin,Sum_transf,fill=High_Low))+
  geom_bar(stat = "identity",width = 0.5)+
  coord_flip()+
  expand_limits(y=0)
  
```

```{r}
datoxin %>% 
  count(origin,High_Low,Sum_transf,sort=TRUE) %>% 
  mutate(origin= fct_reorder(origin,Sum_transf,sum)) %>% 
  ggplot(aes(origin,Sum_transf,fill=High_Low))+
  geom_bar(stat = "identity",width = 0.5)+
  coord_flip()+
  expand_limits(y=0) +
  scale_fill_grey(start = 0.8, end = 0.5 )+
  labs(x="Country of origin",y="Counts with High or Low Sum of T2_HT2 toxin")+
  theme_classic()+
  theme(legend.position =c(0.9,0.2),text=element_text(size=12, family="Times New Roman"))+
  scale_y_continuous(limits = c(0,500))
```


```{r}
datoxin %>% 
  group_by(origin,Period) %>% 
  ggplot(aes(origin,Sum_transf,fill=Period))+
  geom_boxplot()+
  scale_fill_grey(start = 0.2, end = 0.5 )+
  expand_limits(y=0)
```


```{r,eval=FALSE}
datoxin %>% 
  count(origin,High_Low,Sum_transf,sort=TRUE) %>% 
  mutate(origin= fct_reorder(origin,Sum_transf,sum)) %>% 
  ggplot(aes(origin,Sum_transf,fill=High_Low))+
  geom_bar(stat = "identity")+
  coord_flip()+
  expand_limits(y=0)
```

```{r,eval=FALSE}
datoxin %>% 
  count(origin,High_Low,Sum_transf,sort=TRUE) %>% 
   mutate(Perce=n/sum(n)) %>% 
  mutate(origin= fct_reorder(origin,Sum_transf,sum)) %>% 
   ggplot(aes(origin,Perce,fill=High_Low))+
  geom_bar(stat = "identity")+
  coord_flip()+
  expand_limits(y=0)

#table(dat$origin)

```

```{r}
#colgray <- gray.colors(5, start = 0.2, end = 0.9, alpha = 0.1, rev = FALSE)
colgray2 <- gray.colors(10, start = 0.1, end = 0.9, gamma = 5,alpha = 0.2)

a <- datoxin %>%
  count(Year,Sum_transf,sort=TRUE) %>% 
  ggplot(aes(Sum_transf,fill=Year)) + 
  geom_histogram(binwidth=0.15,alpha=0.7)+
  scale_x_continuous(name = "Sum T2-HT2 Toxin") +
  scale_y_continuous(name = "count", limits=c(0,30),labels = c("0", "10", "20", "30")) +
  scale_fill_manual(values = colgray2 ) +
    theme_classic()+
  theme(legend.position = c(0.88,0.8),legend.background = element_blank())


```


```{r}
b <- datoxin %>%
  count(Year,T2_Transf ,sort=TRUE) %>% 
  ggplot(aes(T2_Transf ,fill=Year)) + 
  geom_histogram(binwidth=0.15,alpha=0.7)+
  scale_x_continuous(name = "T2 Toxin") +
  scale_y_continuous(name = "count", limits  = c(0, 30)) +
 scale_fill_manual(values = colgray2 ) +
    theme_classic()+
  theme(legend.position = c(0.88,0.8),legend.background = element_blank()) 

```


```{r}
c <- datoxin %>%
  count(Year,HT2_Trans ,sort=TRUE) %>% 
  ggplot(aes(HT2_Trans ,fill=Year)) + 
  geom_histogram(binwidth=0.15,alpha=0.7)+
  scale_x_continuous(name = "HT2 Toxin") +
  scale_y_continuous(name = "count", limits  = c(0, 30)) +
  scale_fill_manual(values = colgray2 ) +
    theme_classic()+
  theme(legend.position = c(0.88,0.8),legend.background = element_blank())

```







```{r}
d <- datoxin %>%
  count(Year,FungalLOD_tran ,sort=TRUE) %>% 
  ggplot(aes(FungalLOD_tran ,fill=Year)) + 
  geom_histogram(binwidth=0.15,alpha=0.7)+
  scale_x_continuous(name = "Fungal Biomass") +
  scale_y_continuous(name = "count", limits  = c(0, 40)) +
   scale_fill_manual(values = colgray2 ) +
    theme_classic()+
  theme(legend.position = c(0.82,0.84),legend.background = element_blank())
```


```{r}
pdf("Hist_Toxin.pdf",width=8,height=6)
multiplot(a,b,c,d,cols = 2)
dev.off()
```


```{r,eval=FALSE}

datoxin %>% 
  group_by(High_Low,Period) %>% 
  ggplot(aes(High_Low,Sum_transf,fill=Period))+
  geom_boxplot()+
  #scale_fill_grey(start = 0.2, end = 0.5 )+
  expand_limits(y=0)
```



# Other toxins
### DON
```{r}
datoxin %>%
  mutate(logDON=log(0.1+DON)) %>% 
  count(Year,logDON ,sort=TRUE) %>% 
  ggplot(aes(logDON ,fill=Year)) + 
  geom_histogram(binwidth=0.1,alpha=0.3)+
  scale_x_continuous(name = "DON") +
  scale_y_continuous(name = "count") +
  scale_fill_manual(values = c("#D55E00", "#009E73")) +
  theme(legend.position = c(0.9,0.9),legend.background = element_blank())


#datoxin %>%
 # mutate(logDON=log(0.1+DON)) %>% 
#  count(Year,logDON ,sort=TRUE) %>% 
#  ggplot(aes(logDON,fill=Year)) + 
#  geom_histogram(binwidth=0.2,alpha=0.7)+
 # scale_x_continuous(name = "Fungal Biomass") +
#  scale_y_continuous(name = "count", limits  = c(0, 10)) +
#  scale_fill_manual(values = c("#D55E00", "#009E73")) +
 # theme(legend.position = c(0.15,0.8),legend.background = element_blank())

#table(datoxin$DON==0)

#DONcontamination <- datoxin[!datoxin$DON==0,]

#table(DONcontamination$Genebank)
#table(DONcontamination$Period)
#table(DONcontamination$origin)
#which(table(unique(DONcontamination$GID2))>0)
#table(DONcontamination$Year)
#print(DONcontamination, n=30)
#summary(DONcontamination)
```

###### 718 samples did not show value for DON (0). Just 22 samples showed DON>0.-->
###### These were the genotypes G1 G114 G137 G155   G2  G29   G4  G42  G74  G88   G9 
###### More Modern varieties than old varities showed DON infection.


### NIV
```{r}
datoxin %>%
  mutate(logNIV=log(0.1+NIV)) %>% 
  count(Year,logNIV ,sort=TRUE) %>% 
  ggplot(aes(logNIV ,fill=Year)) + 
  geom_histogram(binwidth=0.5,alpha=0.3)+
  scale_x_continuous(name = "NIV") +
  scale_y_continuous(name = "count") +
  scale_fill_manual(values = c("#D55E00", "#009E73")) +
  theme(legend.position = c(0.9,0.9),legend.background = element_blank())
```

### ZEN
```{r}
datoxin %>%
  mutate(logZEN=log(0.1+ZEN)) %>% 
  count(Year,logZEN ,sort=TRUE) %>% 
  ggplot(aes(logZEN ,fill=Year)) + 
  geom_histogram(binwidth=0.1,alpha=0.3)+
  scale_x_continuous(name = "ZEN") +
  scale_y_continuous(name = "count") +
  scale_fill_manual(values = c("#D55E00", "#009E73")) +
  theme(legend.position = c(0.9,0.9),legend.background = element_blank())
```

### AcDON_15
```{r}
datoxin %>%
  mutate(logAcD_15=log(0.1+AcDON_15)) %>% 
  count(Year,logAcD_15 ,sort=TRUE) %>% 
  ggplot(aes(logAcD_15 ,fill=Year)) + 
  geom_histogram(binwidth=0.1,alpha=0.3)+
  scale_x_continuous(name = "AcDON_15") +
  scale_y_continuous(name = "count") +
  scale_fill_manual(values = c("#D55E00", "#009E73")) +
  theme(legend.position = c(0.9,0.9),legend.background = element_blank())
```


# Summary Pheno data for T2-HT2

```{r}
all <- datoxin %>% 
  summarize(MeanSum=mean(Sum_transf), MeanHT2=mean(HT2_Trans),MeanT2=mean(T2_Transf),MeanFungal=mean(FungalLOD_tran), n=n(),
            sdSum=sd(Sum_transf),sdHT=sd(HT2_Trans), sdT2=sd(T2_Transf),sdFun=sd(FungalLOD_tran), 
            seSum=sdSum/sqrt(n),seHT2=sdHT/sqrt(n),seT2=sdT2/sqrt(n), seFun=sdFun/sqrt(n))


all %>%
kable(digits = 2)%>%
  kable_styling(bootstrap_options = "striped", full_width = F,font_size = 12,fixed_thead = T)  
 

```


#### By Year
```{r}
t1 <- datoxin %>% 
  group_by(Year) %>% 
  summarize(MeanSum=mean(Sum_transf), MeanHT2=mean(HT2_Trans),MeanT2=mean(T2_Transf),MeanFungal=mean(FungalLOD_tran), n=n(),
            sdSum=sd(Sum_transf),sdHT=sd(HT2_Trans), sdT2=sd(T2_Transf),sdFun=sd(FungalLOD_tran), 
            seSum=sdSum/sqrt(n),seHT2=sdHT/sqrt(n),seT2=sdT2/sqrt(n), seFun=sdFun/sqrt(n))


t1 %>%
kable(digits = 2)%>%
  kable_styling(bootstrap_options = "striped", full_width = F,font_size = 12,fixed_thead = T)  
 

```

#### By Origin
```{r}
t2 <- datoxin %>% 
  group_by(origin) %>% 
  summarize(MeanSum=mean(Sum_transf), MeanHT2=mean(HT2_Trans),MeanT2=mean(T2_Transf),MeanFungal=mean(FungalLOD_tran), n=n(),
            sdSum=sd(Sum_transf),sdHT=sd(HT2_Trans), sdT2=sd(T2_Transf),sdFun=sd(FungalLOD_tran), 
            seSum=sdSum/sqrt(n),seHT2=sdHT/sqrt(n),seT2=sdT2/sqrt(n), seFun=sdFun/sqrt(n))      
t2 %>%
kable(digits = 2)%>%
  kable_styling(bootstrap_options = "striped", full_width = F,font_size = 12,fixed_thead = T)  
 

```

#### By Period

```{r}
t3 <- datoxin %>% 
  group_by(Period) %>% 
  summarize(MeanSum=mean(Sum_transf), MeanHT2=mean(HT2_Trans),MeanT2=mean(T2_Transf),MeanFungal=mean(FungalLOD_tran), n=n(),
            sdSum=sd(Sum_transf),sdHT=sd(HT2_Trans), sdT2=sd(T2_Transf),sdFun=sd(FungalLOD_tran), 
            seSum=sdSum/sqrt(n),seHT2=sdHT/sqrt(n),seT2=sdT2/sqrt(n), seFun=sdFun/sqrt(n))

t3 %>%
kable(digits = 2)%>%
  kable_styling(bootstrap_options = "striped", full_width = F,font_size = 12,fixed_thead = T)
 

```
#### By Year and Period

```{r}
t4 <- datoxin %>% 
  group_by(Year,origin) %>% 
  summarize(MeanSum=mean(Sum_transf), MeanHT2=mean(HT2_Trans),MeanT2=mean(T2_Transf),MeanFungal=mean(FungalLOD_tran), n=n(),
            sdSum=sd(Sum_transf),sdHT=sd(HT2_Trans), sdT2=sd(T2_Transf),sdFun=sd(FungalLOD_tran), 
            seSum=sdSum/sqrt(n),seHT2=sdHT/sqrt(n),seT2=sdT2/sqrt(n), seFun=sdFun/sqrt(n))

t4 %>%
kable(digits = 2)%>%
  kable_styling(bootstrap_options = "striped", full_width = F,font_size = 12,fixed_thead = T)
 
```

#### By Genebank
```{r}
t5 <- datoxin %>% 
  group_by(Genebank) %>% 
  summarize(MeanSum=mean(Sum_transf), MeanHT2=mean(HT2_Trans),MeanT2=mean(T2_Transf),MeanFungal=mean(FungalLOD_tran), n=n(),
            sdSum=sd(Sum_transf),sdHT=sd(HT2_Trans), sdT2=sd(T2_Transf),sdFun=sd(FungalLOD_tran), 
            seSum=sdSum/sqrt(n),seHT2=sdHT/sqrt(n),seT2=sdT2/sqrt(n), seFun=sdFun/sqrt(n))

t5 %>%
kable(digits = 2)%>%
  kable_styling(bootstrap_options = "striped", full_width = F,font_size = 12,fixed_thead = T) 
 
```


#### By Block
```{r}

t6 <- datoxin %>% 
  group_by(Block) %>% 
  summarize(MeanSum=mean(Sum_transf), MeanHT2=mean(HT2_Trans),MeanT2=mean(T2_Transf),MeanFungal=mean(FungalLOD_tran), n=n(),
            sdSum=sd(Sum_transf),sdHT=sd(HT2_Trans), sdT2=sd(T2_Transf),sdFun=sd(FungalLOD_tran), 
            seSum=sdSum/sqrt(n),seHT2=sdHT/sqrt(n),seT2=sdT2/sqrt(n), seFun=sdFun/sqrt(n))

t6 %>%
kable(digits = 2)%>%
  kable_styling(bootstrap_options = "striped", full_width = F,font_size = 12,fixed_thead = T)
 
```

# Summary phenotypes other toxins

#### By year
```{r}

t7 <- datoxin %>% 
  group_by(Year) %>% 
  summarize(MeanZen=mean(ZEN), MeanNIV=mean(NIV),MeanDON=mean(DON),MeanAcD_15=mean(AcDON_15), n=n(),
            sdZen=sd(ZEN),sdNIV=sd(NIV), sdDON=sd(DON),sdAcD_15=sd(AcDON_15), 
            seZen=sdZen/sqrt(n),seNIV=sdNIV/sqrt(n),seDON=sdDON/sqrt(n), seAcD_15=sdAcD_15/sqrt(n))
t7 %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = "striped", full_width = F,font_size = 12,fixed_thead = T)

```
#### By Period

```{r}

t8 <- datoxin %>% 
  group_by(Period) %>% 
  summarize(MeanZen=mean(ZEN), MeanNIV=mean(NIV),MeanDON=mean(DON),MeanAcD_15=mean(AcDON_15), n=n(),
            sdZen=sd(ZEN),sdNIV=sd(NIV), sdDON=sd(DON),sdAcD_15=sd(AcDON_15), 
            seZen=sdZen/sqrt(n),seNIV=sdNIV/sqrt(n),seDON=sdDON/sqrt(n), seAcD_15=sdAcD_15/sqrt(n))
t8 %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = "striped", full_width = F,font_size = 12,fixed_thead = T)

```

#### By Origin
```{r}
t9 <- datoxin %>% 
  group_by(origin) %>% 
  summarize(MeanZen=mean(ZEN), MeanNIV=mean(NIV),MeanDON=mean(DON),MeanAcD_15=mean(AcDON_15), n=n(),
            sdZen=sd(ZEN),sdNIV=sd(NIV), sdDON=sd(DON),sdAcD_15=sd(AcDON_15), 
            seZen=sdZen/sqrt(n),seNIV=sdNIV/sqrt(n),seDON=sdDON/sqrt(n), seAcD_15=sdAcD_15/sqrt(n))
t9 %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = "striped", full_width = F,font_size = 12,fixed_thead = T)
```



#### By Year and Origin

```{r}
t10 <- datoxin %>% 
  group_by(Year,origin) %>% 
  summarize(MeanZen=mean(ZEN), MeanNIV=mean(NIV),MeanDON=mean(DON),MeanAcD_15=mean(AcDON_15), n=n(),
            sdZen=sd(ZEN),sdNIV=sd(NIV), sdDON=sd(DON),sdAcD_15=sd(AcDON_15), 
            seZen=sdZen/sqrt(n),seNIV=sdNIV/sqrt(n),seDON=sdDON/sqrt(n), seAcD_15=sdAcD_15/sqrt(n))
t10 %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = "striped", full_width = F,font_size = 12,fixed_thead = T)
```


#### By Year and Period

```{r}
t11 <- datoxin %>% 
  group_by(Year, Period) %>% 
  summarize(MeanZen=mean(ZEN), MeanNIV=mean(NIV),MeanDON=mean(DON),MeanAcD_15=mean(AcDON_15), n=n(),
            sdZen=sd(ZEN),sdNIV=sd(NIV), sdDON=sd(DON),sdAcD_15=sd(AcDON_15), 
            seZen=sdZen/sqrt(n),seNIV=sdNIV/sqrt(n),seDON=sdDON/sqrt(n), seAcD_15=sdAcD_15/sqrt(n))
t11 %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = "striped", full_width = F,font_size = 12,fixed_thead = T)
```



#### By Genebank

```{r}
t12 <- datoxin %>% 
  group_by(Genebank) %>% 
  summarize(MeanZen=mean(ZEN), MeanNIV=mean(NIV),MeanDON=mean(DON),MeanAcD_15=mean(AcDON_15), n=n(),
            sdZen=sd(ZEN),sdNIV=sd(NIV), sdDON=sd(DON),sdAcD_15=sd(AcDON_15), 
            seZen=sdZen/sqrt(n),seNIV=sdNIV/sqrt(n),seDON=sdDON/sqrt(n), seAcD_15=sdAcD_15/sqrt(n))
t12 %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = "striped", full_width = F,font_size = 12,fixed_thead = T)
```

#### By Block

```{r}

t13 <- datoxin %>% 
  group_by(Block) %>% 
  summarize(MeanZen=mean(ZEN), MeanNIV=mean(NIV),MeanDON=mean(DON),MeanAcD_15=mean(AcDON_15), n=n(),
            sdZen=sd(ZEN),sdNIV=sd(NIV), sdDON=sd(DON),sdAcD_15=sd(AcDON_15), 
            seZen=sdZen/sqrt(n),seNIV=sdNIV/sqrt(n),seDON=sdDON/sqrt(n), seAcD_15=sdAcD_15/sqrt(n))
t13 %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = "striped", full_width = F,font_size = 12,fixed_thead = T)

```


```{r,echo=FALSE,results='hide'}
library(lme4)
library(insight) #Calculate variance components easily
library(sommer)
m1 <-lmer(Sum_transf ~ Year+ Block%in%Year+origin+(1|GID2),data=datoxin)
ranefm1<-ranef(m1)
summary(m1)
#get_variance(m1)
Vfix <- get_variance_fixed(m1)
Vres <- get_variance_residual(m1)
Vran <- get_variance_random(m1)
Vdata <- var(datoxin$Sum_transf)
#the sum of the variance components is close to the total variance of the data
Vfix+Vres+Vran

her1 <- round(as.numeric(Vran/(Vran+Vres)),2)

### ANOTHER WAY TO CALCULATE VARIANCES
#Var_Random_effect <- as.numeric(VarCorr(model1)) 
#Var_Residual <- attr(VarCorr(model1), "sc")^2
#Var_Fix_effect <- var(predict(lm(Sum_transf~Year+Year+ Block%in%Year+origin,data=datoxin)))
#Var_Random_effect+Var_Residual+Var_Fix_effect
#the sum of the variance components is close to the total variance of the data
#var(datoxin$Sum_transf)
BLUPS_sum_m1<-data.frame(GID=rownames(ranefm1$GID), Blups=ranefm1$GID$`(Intercept)`)
#hist(BLUPS_sum_m1$Blups,breaks=10)

## MODEL 2 just with YEAR and BLOCK nested to YEAR as fixed effect
m2 <- lmer(Sum_transf ~ Year+ Block%in%Year+(1|GID2),data=datoxin)

m2  <-lmer(Sum_transf ~ Year+ Block%in%Year+origin+(1|GID2),data=datoxin)
ranefm2<-ranef(m2)
#ranefm2$GID$`(Intercept)`

Vfixm2 <- get_variance_fixed(m2)
Vresm2 <- get_variance_residual(m2)
Vranm2 <- get_variance_random(m2)
her2 <- round(as.numeric(Vranm2/(Vranm2+Vresm2)),2) #divided by 2 reps?
BLUPS_sum_m2<-data.frame(GID=rownames(ranefm2$GID), Blups=ranefm2$GID$`(Intercept)`)
#hist(BLUPS_sum_m2$Blups,breaks=10)


## MODEL 3 USING SOMMER BLOCK nested to YEAR as fixed effect.

m3 <- mmer(Sum_transf~Year+ Block%in%Year+origin, random =~ GID2,getPEV = T,data = datoxin)
m4 <- mmer(T2_Transf~Year+ Block%in%Year+origin, random =~ GID2,getPEV = T,data = datoxin)
m5 <- mmer(HT2_Trans~Year+ Block%in%Year+origin, random =~ GID2,getPEV = T,data = datoxin)
m6 <- mmer(FungalLOD_tran~Year+ Block%in%Year+origin, random =~ GID2,getPEV = T,data = datoxin)

plot(m3)
title(main="SumT2HT2")
plot(m4)
title(main="T2")
plot(m5)
title(main="HT2")
plot(m6)
title(main="Fungal Biomass")

colgray2
hist(residuals.mmer(m3),col="#94949433", main = "SumT2HT2",xlab="Sum T2-HT2")
hist(residuals.mmer(m4),col="#AAAAAA33", main = "T2",xlab="T2")
hist(residuals.mmer(m5),col="#C3C3C333", main = "HT2",xlab="HT2")
hist(residuals.mmer(m6),col="#CCCCCC33", main = "Biomass",xlab="Fungal Biomass")


summary(m3)
pin(m3, Vp.Sum_transf~(V1+V2)) #phenotypic variance conditional on fixed effects in the model
herSUM <- pin(m3, h2.Sum_transf~V1/(V1+V2/2))

BLUPS_SUM <-as.data.frame(m3$U$GID2$Sum_transf)
BLUPS_T2<-as.data.frame(m4$U$GID2$T2_Transf)
BLUPS_HT2<-as.data.frame(m5$U$GID2$HT2_Trans)
BLUPS_Biomass<-as.data.frame(m6$U$GID2$FungalLOD_tran)

BLUPs <- as.data.frame(cbind(BLUPS_SUM,BLUPS_T2,BLUPS_HT2,BLUPS_Biomass))
colnames(BLUPs) <- c("Blups_SUM","Blups_T2","Blups_HT2","Blups_Biomass")

write.csv(BLUPs,file="BLUPs_SumT2_HT2.csv")
herSUM <- pin(m3, h2.Sum_transf~V1/(V1+V2/2))
herT2 <- pin(m4, h2.T2_Transf~V1/(V1+V2/2))
herHT2 <- pin(m5, h2.HT2_Trans~V1/(V1+V2/2))
herBiomass <- pin(m6, h2.FungalLOD_tran~V1/(V1+V2/2))

heritability <- as.data.frame(rbind(herSUM,herT2,herHT2,herBiomass))


dattop10 <- dat[dat$accession_name%in%top10 & dat$Year==2017,]
write.csv(dattop10,file="top10.csv")

top10 <- c("J_geva Koidukaer",
"FLORALE","Nidar I",
"STARLETTE DE MOYENCOURT",
"KENT BERLIE",
"Kwant",
"Wexford Tawny",
"Sandy",
"Urai tf.",
"Glasnevin Major")
write.csv(BLUPs,"Blups.T2.csv")
BLUPs[sort(BLUPs$Blups_SUM,decreasing = F)[1:10]]
#why this doesn´t work
K.mat2 <- K.mat[rownames(K.mat)%in%datoxin$GID2,colnames(K.mat)%in%datoxin$GID2]
#mmer(Sum_transf~Year+ Block%in%Year+origin, random =~ GID2+GID2*Year,G=list(GID2=K.mat2),getPEV = T,data = datoxin)

```

### BLUPS Graphs

## Assumptions
```{r}
plot(m3)
title(main="SumT2HT2")
plot(m4)
title(main="T2")
plot(m5)
title(main="HT2")
plot(m6)
title(main="Fungal Biomass")
```

```{r}

b1 <- BLUPs %>% 
  ggplot(aes(Blups_SUM))+
  geom_histogram(bins=15,alpha=0.4)+
  theme_classic()

b2 <- BLUPs %>% 
  ggplot(aes(Blups_T2))+
  geom_histogram(bins=15,alpha=0.4)+
  theme_classic()

b3 <- BLUPs %>% 
  ggplot(aes(Blups_HT2))+
  geom_histogram(bins=15,alpha=0.4)+
  theme_classic()

b4 <- BLUPs %>% 
  ggplot(aes(Blups_Biomass))+
  geom_histogram(bins=13,alpha=0.4)+
  theme_classic()

multiplot(b1,b2,b3,b4,cols=2)

```

#### BLUPs summary info
```{r}
sumBLUPs <- as.data.frame(cbind(
summary(BLUPs$Blups_SUM),
summary(BLUPs$Blups_T2),
summary(BLUPs$Blups_HT2),
summary(BLUPs$Blups_Biomass)))

colnames(sumBLUPs) <- c("Blups_SUM","Blups_T2","Blups_HT2","Blups_Biomass")

sumBLUPs %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = "striped", full_width = F,font_size = 12,fixed_thead = T)

```

### Heritability info
```{r}
heritability %>% 
  kable(digits = 3) %>% 
  kable_styling(bootstrap_options = "striped", full_width = F,font_size = 12,fixed_thead = T)

```

# Summary Phenotype without transformation
```{r,results='hide'}
str(datoxin)

m7 <- mmer(SumT2HT2~Year+ Block%in%Year+origin, random =~ GID2,getPEV = T,data = datoxin)
m8 <- mmer(T2~Year+ Block%in%Year+origin, random =~ GID2,getPEV = T,data = datoxin)
m9 <- mmer(HT2~Year+ Block%in%Year+origin, random =~ GID2,getPEV = T,data = datoxin)
m10 <- mmer(FungalLOD_tran~Year+ Block%in%Year+origin, random =~ GID2,getPEV = T,data = datoxin)

summary(m7)
pin(m7, Vp.Sum_transf~(V1+V2)) #phenotypic variance conditional on fixed effects in the model
herSUM <- pin(m7, h2.SumT2HT2~V1/(V1+V2/2))

BLUPS_SUM <-as.data.frame(m7$U$GID2$SumT2HT2)
BLUPS_T2<-as.data.frame(m8$U$GID2$T2)
BLUPS_HT2<-as.data.frame(m9$U$GID2$HT2)
BLUPS_Biomass<-as.data.frame(m10$U$GID2$FungalLOD_tran)

BLUPs <- as.data.frame(cbind(BLUPS_SUM,BLUPS_T2,BLUPS_HT2,BLUPS_Biomass))
colnames(BLUPs) <- c("Blups_SUM","Blups_T2","Blups_HT2","Blups_Biomass")

herSUM2 <- pin(m7, h2.Sum_transf~V1/(V1+V2/4))
herT22 <- pin(m8, h2.T2_Transf~V1/(V1+V2/4))
herHT22 <- pin(m9, h2.HT2_Trans~V1/(V1+V2/4))
herBiomass2 <- pin(m10, h2.FungalLOD_tran~V1/(V1+V2/4))

heritability2 <- as.data.frame(rbind(herSUM2,herT22,herHT22,herBiomass2))

summary(datoxin$FungalLOD_tran)
```

```{r}
Percentageinfectionperorigin <- datoxin %>% 
  select(High_Low,origin,Sum_transf) %>% 
  group_by(origin) %>% 
  count(High=High_Low=="High", Low=High_Low=="Low",n()) %>% 
  mutate(per=(n/`n()`)*100) %>% 
  select(origin,per)

PercetageTotalinfection <- datoxin %>% 
  select(High_Low,Sum_transf) %>% 
   count(High=High_Low=="High", Low=High_Low=="Low",n()) %>% 
  mutate(per=(n/`n()`)*100) %>% 
  select(per)
  
```


```{r}

  

toxin <- datoxin %>% 
  mutate(origin=fct_reorder(origin,Sum_transf,sum)) %>% 
  ggplot(aes(origin,Sum_transf,fill=High_Low))+
  geom_bar(stat = "identity",width = 0.5)+
  coord_flip()+
  expand_limits(y=0) +
  scale_fill_grey(start = 0.9, end = 0.5,name = "Sum T-2 & HT-2", labels = c("High values", "Low values"))+
  labs(x="Country of Origin",y="Counts with High or Low Sum of T-2 & HT-2 toxin")+
  theme_classic()+
  theme(legend.position =c(0.9,0.2),text=element_text(size=12, family="Times New Roman"))+
  scale_y_continuous(limits = c(0,600))

toxin+
  annotate("text", x = 1, y = 150, label = "95%",color="white")+
  annotate("text", x = 1, y = 320, label = "5%",size=3.5,color="black")+
  annotate("text", x = 2, y = 150, label = "92%",color="white")+
  annotate("text", x = 2, y = 310, label = "8%",size=3.5,color="black")+
  annotate("text", x = 3, y = 150, label = "92%",color="white")+
  annotate("text", x = 3, y = 322, label = "8%",size=3.5,color="black")+
  annotate("text", x = 4, y = 150, label = "88%",color="white")+
  annotate("text", x = 4, y = 325, label = "12%",size=3.5,color="black")+
  annotate("text", x = 5, y = 150, label = "90%",color="white")+
  annotate("text", x = 5, y = 340, label = "10%",size=3.5,color="black")+
  annotate("text", x = 6, y = 150, label = "81%",color="white")+
  annotate("text", x = 6, y = 330, label = "19%",size=3.5,color="black")+
  annotate("text", x = 7, y = 150, label = "87%",color="white")+
  annotate("text", x = 7, y = 350, label = "13%",size=3.5,color="black")+
  annotate("text", x = 8, y = 150, label = "91%",color="white")+
  annotate("text", x = 8, y = 490, label = "9%",size=3.5,color="black")
  ggsave("Count_HIGH_LOW.png",width=8,height=6, dpi=700)
```
