---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# GEmetrics

<!-- badges: start -->
<!-- badges: end -->

The goal of GEmetrics is to provide functions to calculate the best linear unbiased prediction (BLUP) of the following genotype-by-environment (GE) metrics: ecovalence, environmental variance, Finlay and Wilkinson regression and Lin and Binns superiority measure, based on a multi-environment genomic prediction model.

## Installation

You can install GEmetrics directly from the CRAN:

``` r
install.packages(pkg='GEmetrics',repos='https://cran.r-project.org/')
```

or from GitHub:

``` r
install.packages(pkg='devtools',repos='https://cran.r-project.org/')     ## install devtools
devtools::install_git('https://github.com/TheRocinante-lab/GEmetrics')   ## install GEmetrics from GitHub
```

## Example

This is a basic example which shows you how to calculate the BLUP of GE metrics. 

#### Simulate phenotypic data

Multi-environment trial data is first simulated based on the "wheat" dataset from BGLR. 

A design data frame is generated displaying all combinations of genotypes and environments. 
Some combinations are discarded to set sparseness in the data (75% here).

```{r}
## Set seed for reproductibility
set.seed(123)

## Load "wheat" dataset from BGLR
data("wheat",package = "BGLR")

## Generate a design data frame for all genotypes in 5 environments
Design <- expand.grid(Genotype=rownames(wheat.A),Environment=paste0("Env",1:5))

## Set sparseness by discarding 75% of the combinatons
Design <- Design[-sample(nrow(Design),round(nrow(Design)*3/4)),]
head(Design)
```

Phenotypes are then simulated using trait and environments parameters: 

* h2: heritability, either a single value for a heritability common to all environments (e.g. 0.5 in the example below), or a vector of heritabilities associated with each environment 
* rho: genetic correlations between environment pairs, either a single value for a genetic correlation common to all environment pairs (e.g. 0.5 in the example below), or a square correlation matrix
* sd_mu: standard deviation of the Gaussian distribution in which environment means are drawn (e.g. 1 in the example below)

```{r}
## Simulate phenotypic data with default parameter values
DataSim <- GEmetrics::Simulate_MET_data(Design=Design,K=wheat.A,h2=0.5,rho=0.5,sd_mu=1)
```

The resulting DataSim object include:

(i) Pheno: data frame with simulated phenotypes

```{r}
## Simulated phenotypes
head(DataSim$Pheno)
```

(ii) EnvBV: matrix of simulated environment-specific breeding values

```{r}
## Simulated environment-specific breeding values
head(DataSim$EnvBV)
```

(iii) Omega_G: genetic covariance matrix between environments

```{r}
## Genetic covariance matrix between environments
DataSim$Omega_G
```

(iv) Omega_E: error covariance matrix between environments

```{r}
## Error covariance matrix between environments
DataSim$Omega_E
```

#### Estimate variance components using BGLR

From simulated data, variance components can be estimated using an inference 
method like [BGLR](https://github.com/gdlc/BGLR-R), or any other methods able to 
infer Omega_G and Omega_E. 

First, the phenotypic data frame must be transformed into a phenotypic response matrix.

```{r}
## Generate the phenotypic response matrix for BGLR and the corresponding K matrix
BGLR_data <- GEmetrics::BGLR_format(Pheno=DataSim$Pheno,K=wheat.A)
head(BGLR_data$BGLR_pheno)
```

The inference can be done using the "Multitrait" function of BGLR to estimate the
Omega_G and Omega_E covariance matrices. Note that the current CRAN version of BGLR 
(October 2023) may lead to an issue when the phenotypic data is very sparse, but
not the most recent GitHub version.

```{r}
## Run BGLR inference
ETA<-list(list(K=BGLR_data$BGLR_K,model="RKHS"))
BGLR_results <- BGLR::Multitrait(y=BGLR_data$BGLR_pheno,ETA=ETA,
                                 resCov=list(type="DIAG"),
                                 nIter=1000,burnIn=500,verbose = F,saveAt = "Test_")
unlink(c("Test_R.dat","Test_Omega_1.dat","Test_mu.dat"))
Omega_G <- BGLR_results$ETA[[1]]$Cov$Omega
Omega_E <- BGLR_results$resCov$R
rownames(Omega_E) <- rownames(Omega_E) <- rownames(Omega_G)
```

The estimate of the genetic covariance matrix Omega_G is:

```{r}
Omega_G
```

and the estimate of the error covariance matrix Omega_E is:

```{r}
Omega_E
```


#### Calculate BLUP and conditional variance of environment-specific breeding values

The BLUP and the conditional variance of environment-specific breeding values can 
be calculated from the phenotypes and the variance component estimates. 
Note that the BLUPs could also be obtained directly from BGLR outputs.

```{r}
## Calculate BLUP and conditional variance
BlupEnvBV <- GEmetrics::EnvBV_blup(Pheno=DataSim$Pheno,K=wheat.A,Omega_G=Omega_G,Omega_E=Omega_E)
```

The BLUPs obtained:

```{r}
head(BlupEnvBV$G_hat)
```

and the conditional variance matrix:

```{r}
BlupEnvBV$P[1:5,1:5]
```

#### Obtain GE metrics estimates

Each GE metric can be estimated using the complete BLUP involving both
the squared expectation and the variance term:

```{r}
metrics <- c("Ecovalence","EnvironmentalVar","FinlayWilkRegression","LinBinns")
GEmetrics_hat_geno_exp_var <- sapply(metrics,function(m)
  GEmetrics::GEmetrics_blup(G_hat=BlupEnvBV$G_hat,metric=m,P=BlupEnvBV$P))
head(GEmetrics_hat_geno_exp_var)
```
or using the partial BLUP including the the squared expectation only:

```{r}
metrics <- c("Ecovalence","EnvironmentalVar","FinlayWilkRegression","LinBinns")
GEmetrics_hat_geno_exp <- sapply(metrics,function(m)
  GEmetrics::GEmetrics_blup(G_hat=BlupEnvBV$G_hat,metric=m))
head(GEmetrics_hat_geno_exp)
```

Estimated can be compared to the true GE metric values obtained from simulated environment-specific 
breeding values using the correlation:

```{r}
GEmetrics_true <- sapply(metrics,function(m)GEmetrics::GEmetrics_blup(G_hat=DataSim$EnvBV,metric=m,P=NULL))
data.frame("Geno_Exp_Var" = diag(cor(GEmetrics_hat_geno_exp_var,GEmetrics_true)),
           "Geno_Exp" = diag(cor(GEmetrics_hat_geno_exp,GEmetrics_true)))
```

or the root mean-square error of estimation:

```{r}
data.frame("Geno_Exp_Var" = sqrt(colMeans((GEmetrics_hat_geno_exp_var-GEmetrics_true)^2)),
           "Geno_Exp" = sqrt(colMeans((GEmetrics_hat_geno_exp-GEmetrics_true)^2)))

```
